---
title: "eds-221 ICE AND METEORS RAHH"
format: html
editor: visual
---
```{r}
library(tidyverse)
library(here)
library(janitor)
```
```{r}
ice <- read_csv(here("data", "ice.csv"), na = "-999")
meteorological <- read_csv(here("data", "meteor.csv"))
```


```{r}
sum_ice <- ice |> group_by(lakeid, year4) |> 
  summarise(mean_ice_duration = mean(ice_duration))
```

```{r}
ggplot(ice, aes(x = lakeid, y = ice_duration)) +
  geom_jitter(size = .5, aes(color = lakeid)) +
  geom_boxplot(alpha= .1, outlier.shape = NA) +
  theme_light() +
  labs(x = "Lake",
       y = "Ice Duration (Days)",
       color = "Lake ID")  
```
The overall trends of this graph shows that Wisconsin Lake had the highest overall year average ice duration in days. Mendota Lake had the lowest ice day year. 

```{r}
ggplot(sum_ice, aes(x = year4, y = mean_ice_duration)) +
  geom_point() +
  theme_classic() +
  labs(x = "Year",
       y = "Mean Ice Duration (days)") +
   geom_smooth(method = "lm", se = FALSE)
```
The average days of ice per year is gradually decreasing over time. 

```{r}

winter <- meteorological |> 
  filter(month %in% c(1,2, 12))

lake_madison_air <- winter |> 
  group_by(year4) |> 
 summarise(ave_air_temp = mean(ave_air_temp_adjusted)) |> 
  ungroup()

ggplot(lake_madison_air, aes(x = year4, y = ave_air_temp)) +
  geom_point() +
  theme_classic() +
  labs(x = "Year",
       y = "Average Air Temperature") +
   geom_smooth(method = "lm", se = FALSE)
```
Air temperature is gradually increasing every year in the months December, January, and February for the Lake Madison Area.

```{r}
join_lake_madison <- left_join(sum_ice, lake_madison_air) 
  
ggplot(join_lake_madison, aes(x = ave_air_temp, y = mean_ice_duration)) +
  geom_point() +
  theme_classic() +
  labs(x = "Average Air Temp (Fahrenheit)",
       y = "Ice Duration (Days)") +
   geom_smooth(method = "lm", se = FALSE)

```
Ice days are gradually decreasing as air temperature is increasing. 


```{r}
thaw_dates_1970 <- ice |> 
  mutate(ice_year = year(ymd(ice_off))) |> 
  mutate(ice_day = yday(ymd(ice_off))) |> 
  filter(ice_year >= 1970)
``` 

```{r}
ggplot(thaw_dates_1970, aes(x = ice_year, y= ice_day)) +
  geom_col() +
  theme_classic() +
  labs(x = "Year",
       y = "Days",
       title = "Lake Madison Ice Melt Dates per Year") +
   geom_smooth(method = "lm", se = FALSE)
```





